---
title: "Wellbeing and Connectedness After Ayahuasca: Findings from a Longtitudinal Pilot Study Among Retreat Participants in Petrópolis, Brazil"
abstract: "coming up next! goat 3"
author:
  - name: "Victor Souza Eduardo"
    affiliation: "CUNY Graduate School of Public Health"
  - name: "Michèle Anne Barocchi, PhD"
    degrees: PhD
    affiliation: "MAPS Italia"
date: last-modified
date-format: medium
last-modified: true
published-title: "Last Updated"
format:
  html:
    output-file: index.html
    toc: true
    toc-depth: 4
    toc-expand: 2
    self-contained: true
    include-in-header: favicon.html
    include-before-body: title.html
    date-block:
      published: false
      modified: true
    html-format:
      enable-download: true
      other-formats: false
    respect-user-color-scheme: true
theme:
  light: arrow-light
  dark: darkly
always_allow_html: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# ---- Libraries ----
library(tidyverse)
library(readxl)
library(table1)
library(shiny)
library(scales)
library(glue)
library(plotly)
library(broom)
library(skimr)
library(kableExtra)
library(stringr)
library(ggthemes)
library(tidytext)
library(textdata)
library(cluster)
library(stringr)
library(ggplot2)
library(emmeans)
library(lme4)
library(readxl)
library(sysfonts)
library(showtext)
library(beepr)
# 
# # ---- Set universal fonts ----
# font_add_google("Lato", "lato")
# showtext_auto()
# 
# # Set Lato as the default ggplot2 font
# theme_set(theme_minimal(base_family = "lato"))
# update_geom_defaults("text",   list(family = "lato"))
# update_geom_defaults("label",  list(family = "lato"))
# update_geom_defaults("point",  list(family = "lato"))

# ---- Basev Import and Clean ----
basev <- read_excel(
  "~/Documents/Akasha/baseline-values 6-30-2025.xlsx",
  col_types = c(
    "skip", # Start Date
    "skip", # End Date
    "skip", # Response Type
    "skip", # IP Address
    "skip", # Progress
    "skip", # Duration (in seconds)
    "numeric", # Finished
    "date", # Recorded Date
    "text", # Response ID
    "skip", # Recipient Last Name
    "skip", # Recipient First Name
    "skip", # Recipient Email
    "skip", # External Data Reference
    "skip", # Location Latitude
    "skip", # Location Longitude
    "skip", # Distribution Channel
    "skip", # User Language
    "skip", # consent
    "skip", # retreat - Selected Choice
    "skip", # retreat - Not displayed - Text
    "numeric", # age
    "skip", # gender - Selected Choice
    "skip", # gender - Not listed - Text
    "text", # Text (country of origin)
    "text", # Text (country of residence)
    "skip", # education
    "skip", # language - Selected Choice
    "skip", # language - Not listed - Text
    "text", # Text (religion or faith)
    "skip", # practices - Selected Choice
    "skip", # practices - Not listed (please specify) - Text
    rep("numeric", 15), # wellbeing items
    "skip", # previous ayahuasca
    "skip", # intention - Selected Choice
    "skip", # intention - Mental health (e.g., depression, anxiety, addiction) - Text
    "skip", # intention - Physical health (e.g., chronic pain) - Text
    "skip", # intention - Not listed - Text
    rep("numeric", 11), # connectedness items
    "text", # connection definition
    "skip", # IMPORTANT: Before clicking Submit… Participant ID: [Field-…]
    "skip", # Score
    "text" # Participant ID
  ),
  skip = 1 # skip the extra row in Qualtrics
) %>%
  rename(
    bl_recorded_date = "Recorded Date",
    bl_response_id = "Response ID",
    country_origin = "Text...5",
    country_residence = "Text...6",
    religion = "Text...7",
    bl_conn_def = "connection definition",
    pid = "Participant ID",
    `bl_wb_1 - I have felt calm and at peace.` =
      "baseline wellbeing - I have felt calm and at peace.",
    `bl_wb_2 - I have felt irritable or easily frustrated.` =
      "baseline wellbeing - I have felt irritable or easily frustrated.",
    `bl_wb_3 - I have experienced feelings of loneliness.` =
      "baseline wellbeing - I have experienced feelings of loneliness.",
    `bl_wb_4 - I have felt down, depressed, or hopeless.` =
      "baseline wellbeing - I have felt down, depressed, or hopeless.",
    `bl_wb_5 - I have felt anxious or \"on edge\".` =
      "baseline wellbeing - I have felt anxious or \"on edge\".",
    `bl_wb_6 - I have experienced physical pain or discomfort.` =
      "baseline wellbeing - I have experienced physical pain or discomfort.",
    `bl_wb_7 - I have had difficulty concentrating or focusing.` =
      "baseline wellbeing - I have had difficulty concentrating or focusing.",
    `bl_wb_8 - I have had difficulty sleeping.` =
      "baseline wellbeing - I have had difficulty sleeping.",
    `bl_wb_9 - I have found it easy to make decisions.` =
      "baseline wellbeing - I have found it easy to make decisions.",
    `bl_wb_10 - I have felt motivated to accomplish daily tasks.` =
      "baseline wellbeing - I have felt motivated to accomplish daily tasks.",
    `bl_wb_11 - I have felt a sense of purpose or meaning in life.` =
      "baseline wellbeing - I have felt a sense of purpose or meaning in life.",
    `bl_wb_12 - I have experienced moments of awe or transcendence.` =
      "baseline wellbeing - I have experienced moments of awe or transcendence.",
    `bl_wb_13 - I have been overall satisfied with my life.` =
      "baseline wellbeing - I have been overall satisfied with my life.",
    `bl_wb_14 - I have felt supported by my social network.` =
      "baseline wellbeing - I have felt supported by my social network.",
    `bl_wb_15 - I have experienced conflict or tension in my relationships.` =
      "baseline wellbeing - I have experienced conflict or tension in my relationships.",
    `bl_nsc_1 - I feel emotionally connected to the natural world.` =
      "connectedness - I feel emotionally connected to the natural world.",
    `bl_nsc_2 - I view humans as separate from nature.` =
      "connectedness - I view humans as separate from nature.",
    `bl_nsc_3 - I notice and appreciate the small details of nature.` =
      "connectedness - I notice and appreciate the small details of nature (e.g., plants, animals, weather).",
    `bl_nsc_4 - I feel a sense of awe and wonder when in nature.` =
      "connectedness - I feel a sense of awe and wonder when in nature.",
    `bl_nsc_5 - I feel motivated to make choices that are more environmentally friendly.` =
      "connectedness - I feel motivated to make choices that are more environmentally friendly (e.g., recycling, conserving energy).",
    `bl_nsc_6 - I feel a strong spiritual bond with the Earth and its living beings.` =
      "connectedness - I feel a strong spiritual bond with the Earth and its living beings.",
    `bl_nsc_7 - I feel connected to others.` =
      "connectedness - I feel connected to others.",
    `bl_nsc_8 - I feel disconnected from others.` =
      "connectedness - I feel disconnected from others.",
    `bl_nsc_9 - I feel like I belong.` =
      "connectedness - I feel like I belong.",
    `bl_nsc_10 - I feel understood by others.` =
      "connectedness - I feel understood by others.",
    `bl_nsc_11 - I feel connected to my community.` =
      "connectedness - I feel connected to my community."
  ) %>%
  mutate(
    bl_recorded_date_r = as.Date(bl_recorded_date),
    retreat = case_when(
      bl_recorded_date_r <= as.Date("2025-02-07") ~ "Ninawa",
      bl_recorded_date_r <= as.Date("2025-02-14") ~ "Colibri Festival",
      bl_recorded_date_r <= as.Date("2025-03-02") ~ "Tumã",
      bl_recorded_date_r <= as.Date("2025-04-25") ~ "Nixiwaka",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(Finished != 0) %>%
  mutate(
    timepoint = "Baseline"
  ) %>%
  relocate(c(pid, bl_response_id, retreat), .before = 1) %>%
  select(-Finished, -bl_recorded_date_r)

# ---- Basetxt Import and Clean ----
basetxt <- read_excel(
  "~/Documents/Akasha/baseline-labels 6-30-2025.xlsx",
  skip = 1,
  col_types = c(
    "skip", # Start Date
    "skip", # End Date
    "skip", # Response Type
    "skip", # IP Address
    "skip", # Progress
    "skip", # Duration (in seconds)
    "skip", # Finished
    "skip", # Recorded Date
    "text", # Response ID
    "skip", # Recipient Last Name
    "skip", # Recipient First Name
    "skip", # Recipient Email
    "skip", # External Data Reference
    "skip", # Location Latitude
    "skip", # Location Longitude
    "skip", # Distribution Channel
    "skip", # User Language
    "skip", # consent
    "skip", # retreat - Selected Choice
    "skip", # retreat - Not displayed - Text
    "skip", # age
    "text", # gender - Selected Choice
    "text", # gender - Not listed - Text
    "skip", # Text (country of origin)
    "skip", # Text (country of residence)
    "text", # education
    "text", # language - Selected Choice
    "text", # language - Not listed - Text
    "skip", # What is your religion or belief system?
    "text", # practices - Selected Choice
    "text", # practices - Not listed (please specify) - Text
    rep("skip", 15), # skip the wellbeing items
    "text", # previous ayahuasca
    "text", # intention - Selected Choice
    "text", # intention - Mental health (e.g., depression, anxiety, addiction) - Text
    "text", # intention - Physical health (e.g., chronic pain) - Text
    "text", # intention - Not listed - Text
    rep("skip", 11), # connectedness items
    "skip", # connection definition
    "skip", # IMPORTANT: Before clicking Submit… Participant ID: [Field-…]
    "skip", # Score
    "skip" # Participant ID
  )
) %>%
  rename(
    prev_ayahuasca = "previous ayahuasca",
    intention = "intention - Selected Choice",
    intent_mental = "intention - Mental health (e.g., depression, anxiety, addiction) - Text",
    intent_physical = "intention - Physical health (e.g., chronic pain) - Text",
    intent_other = "intention - Not listed - Text",
    lang_sel = "language - Selected Choice",
    lang_txt = "language - Not listed - Text",
    gender = "gender - Selected Choice",
    gender_other = "gender - Not listed - Text",
    practices = "practices - Selected Choice",
    practices_other = "practices - Not listed (please specify) - Text"
  ) %>%
  mutate(
    # if they chose “Not listed”, language_not_listed will be non-NA,
    # otherwise it’ll be NA – so coalesce will pick the custom text first
    lang_txt = na_if(lang_txt, ""),
    language = coalesce(lang_txt, lang_sel),
    gender_other = na_if(gender_other, ""),
    gender = coalesce(gender_other, gender),
    practices_other = na_if(practices_other, ""),
    practices = coalesce(practices_other, practices)
  ) %>%
  select(-lang_sel, -lang_txt, -gender_other, -practices_other) %>%
  relocate(`Response ID`, .before = 1)

# ---- Follow-up Import and Clean ---- ----
fu <- read_excel(
  "~/Documents/Akasha/follow-up 6-30-2025.xlsx",
  col_types = c(
    "skip", # Start Date
    "skip", # End Date
    "skip", # Response Type
    "skip", # IP Address
    "skip", # Progress
    "skip", # Duration (in seconds)
    "numeric", # Finished
    "date", # Recorded Date
    "text", # Response ID
    "skip", # Recipient Last Name
    "skip", # Recipient First Name
    "skip", # Recipient Email
    "skip", # External Data Reference
    "skip", # Location Latitude
    "skip", # Location Longitude
    "skip", # Distribution Channel
    "skip", # User Language
    "skip", # consent
    "skip", # What is the Participant ID you were given at the end of your first questionnaire? - Selected Choice
    "text", # What is the Participant ID you were given at the end of your first questionnaire? - My Participant ID is... - Text
    "skip", # age
    "skip", # gender - Selected Choice
    "skip", # gender - Not listed - Text
    "skip", # Country of Origin
    "skip", # Country of Residence
    "skip", # education
    "skip", # language - Selected Choice
    "skip", # language - Not listed - Text
    "skip", # Religion
    "skip", # practices - Selected Choice
    "skip", # practices - Not listed (please specify) - Text
    "numeric", # after ceremony feeling - Relaxed and at peace
    "numeric", # after ceremony feeling - Connected to myself
    "numeric", # after ceremony feeling - Connected to others
    "numeric", # after ceremony feeling - Emotionally balanced
    "numeric", # after ceremony feeling - Insightful about my life or challenges
    "numeric", # after ceremony feeling - Confused and disoriented
    "text", # helpful aspects of ceremony
    "text", # greatest emotion
    "numeric", # ceremonial setting aspects - Presence of Nixiwaka
    "numeric", # ceremonial setting aspects - Presence of Putanny
    "numeric", # ceremonial setting aspects - Rituals and prayers (e.g., group healing, smudging, tobacco offerings)
    "numeric", # ceremonial setting aspects - Group setting (versus individual/alone)
    "numeric", # ceremonial setting aspects - Music (e.g., ícaros, drumming)
    "numeric", # ceremonial setting aspects - Physical environment (e.g., nature, ceremonial room)
    "numeric", # ceremonial setting aspects - Sharing circle insights
    "numeric", # ceremonial setting aspects - Immediate neighbor(s)
    "numeric", # ceremony enhanced ayahuasca
    "numeric", # In my opinion, similar effects could be achieved without a ceremonial setting.
    "text", # ceremonial setting one word
    "numeric", # connectedness - I feel emotionally connected to the natural world.
    "numeric", # connectedness - I view humans as separate from nature.
    "numeric", # connectedness - I notice and appreciate the small details of nature (e.g., plants, animals, weather).
    "numeric", # connectedness - I feel a sense of awe and wonder when in nature.
    "numeric", # connectedness - I feel motivated to make choices that are more environmentally friendly (e.g., recycling, conserving energy).
    "numeric", # connectedness - I feel a strong spiritual bond with the Earth and its living beings.
    "numeric", # connectedness - I feel connected to others.
    "numeric", # connectedness - I feel disconnected from others.
    "numeric", # connectedness - I feel like I belong.
    "numeric", # connectedness - I feel understood by others.
    "numeric", # connectedness - I feel connected to my community.
    "text", # connection definition
    "skip", # IMPORTANT: Before clicking Submit… Participant ID: [Field-…]
    "skip", # Score
    "text" # Participant ID
  ),
  skip = 1
) %>%
  mutate(
    timepoint = "Follow-up"
  ) %>%
  rename(
    pid_auto = "Participant ID",
    pid_manual = "What is the Participant ID you were given at the end of your first questionnaire? - My Participant ID is... - Text",
    fu_conn_def_txt = "connection definition",
    fu_recorded_date = `Recorded Date`,
    fu_response_id = `Response ID`,
    fu_ceremony_enhanced_ayahuasca = `ceremony enhanced ayahuasca`,
    fu_ceremonial_setting_one_word_txt = `ceremonial setting one word`,
    fu_of_1_relaxed_and_at_peace = `after ceremony feeling - Relaxed and at peace`,
    fu_of_2_connected_to_self = `after ceremony feeling - Connected to myself`,
    fu_of_3_connected_to_others = `after ceremony feeling - Connected to others`,
    fu_of_4_emotionally_balanced = `after ceremony feeling - Emotionally balanced`,
    fu_of_5_insightful = `after ceremony feeling - Insightful about my life or challenges`,
    fu_of_6_confused = `after ceremony feeling - Confused and disoriented`,
    fu_similar_effects = `In my opinion, similar effects could be achieved without a ceremonial setting.`,
    fu_helpful_aspects_txt = `helpful aspects of ceremony`,
    fu_greatest_emotion_txt = `greatest emotion`,
    `fu_nsc_1 - I feel emotionally connected to the natural world.` =
      "connectedness - I feel emotionally connected to the natural world.",
    `fu_nsc_2 - I view humans as separate from nature.` =
      "connectedness - I view humans as separate from nature.",
    `fu_nsc_3 - I notice and appreciate the small details of nature.` =
      "connectedness - I notice and appreciate the small details of nature (e.g., plants, animals, weather).",
    `fu_nsc_4 - I feel a sense of awe and wonder when in nature.` =
      "connectedness - I feel a sense of awe and wonder when in nature.",
    `fu_nsc_5 - I feel motivated to make choices that are more environmentally friendly.` =
      "connectedness - I feel motivated to make choices that are more environmentally friendly (e.g., recycling, conserving energy).",
    `fu_nsc_6 - I feel a strong spiritual bond with the Earth and its living beings.` =
      "connectedness - I feel a strong spiritual bond with the Earth and its living beings.",
    `fu_nsc_7 - I feel connected to others.` =
      "connectedness - I feel connected to others.",
    `fu_nsc_8 - I feel disconnected from others.` =
      "connectedness - I feel disconnected from others.",
    `fu_nsc_9 - I feel like I belong.` =
      "connectedness - I feel like I belong.",
    `fu_nsc_10 - I feel understood by others.` =
      "connectedness - I feel understood by others.",
    `fu_nsc_11 - I feel connected to my community.` =
      "connectedness - I feel connected to my community.",
    fu_setting_Presence_Shaman = `ceremonial setting aspects - Presence of Nixiwaka`,
    fu_setting_Presence_Putany = `ceremonial setting aspects - Presence of Putanny`,
    fu_setting_Rituals_Prayers = `ceremonial setting aspects - Rituals and prayers (e.g., group healing, smudging, tobacco offerings)`,
    fu_setting_Group_Setting = `ceremonial setting aspects - Group setting (versus individual/alone)`,
    fu_setting_Music = `ceremonial setting aspects - Music (e.g., ícaros, drumming)`,
    fu_setting_Physical_Environment = `ceremonial setting aspects - Physical environment (e.g., nature, ceremonial room)`,
    fu_setting_Sharing_Circle = `ceremonial setting aspects - Sharing circle insights`,
    fu_setting_Immediate_Neighbors = `ceremonial setting aspects - Immediate neighbor(s)`
  ) %>%
  relocate(c(pid_auto, pid_manual, fu_response_id, fu_recorded_date, timepoint), .before = 1) %>%
  filter(Finished != 0) %>%
  select(-Finished)

# ---- One-month Import and Clean ---- ----

omo <- read_excel("~/Documents/Akasha/one-month 6-30-2025.xlsx",
  col_types = c(
    "skip", # Start Date
    "skip", # End Date
    "skip", # Response Type
    "skip", # IP Address
    "skip", # Progress
    "skip", # Duration (in seconds)
    "numeric", # Finished
    "date", # Recorded Date
    "text", # Response ID
    "skip", # Recipient Last Name
    "skip", # Recipient First Name
    "skip", # Recipient Email
    "skip", # External Data Reference
    "skip", # Location Latitude
    "skip", # Location Longitude
    "skip", # Distribution Channel
    "skip", # User Language
    "skip", # consent
    "skip", # What is the Participant ID you were given at the end of your first questionnaire? - Selected Choice
    "text", # What is the Participant ID you were given at the end of your first questionnaire? - My Participant ID is... - Text
    "skip", # age
    "skip", # gender - Selected Choice
    "skip", # gender - Not listed - Text
    "skip", # Text (country of origin)
    "skip", # Text (country of residence)
    "skip", # education
    "skip", # language - Selected Choice
    "skip", # language - Not listed - Text
    "skip", # Text (any other language)
    "skip", # practices - Selected Choice
    "skip", # practices - Not listed (please specify) - Text
    "numeric", # overall feeling - Relaxed and at peace
    "numeric", # overall feeling - Connected to myself
    "numeric", # overall feeling - Connected to others
    "numeric", # overall feeling - Emotionally balanced
    "numeric", # overall feeling - Insightful about my life or challenges
    "numeric", # overall feeling - Confused and disoriented
    "text", # greatest emotion
    "numeric", # connectedness - I feel emotionally connected to the natural world.
    "numeric", # connectedness - I view humans as separate from nature.
    "numeric", # connectedness - I notice and appreciate the small details of nature (e.g., plants, animals, weather).
    "numeric", # connectedness - I feel a sense of awe and wonder when in nature.
    "numeric", # connectedness - I feel motivated to make choices that are more environmentally friendly (e.g., recycling, conserving energy).
    "numeric", # connectedness - I feel a strong spiritual bond with the Earth and its living beings.
    "numeric", # connectedness - I feel connected to others.
    "numeric", # connectedness - I feel disconnected from others.
    "numeric", # connectedness - I feel like I belong.
    "numeric", # connectedness - I feel understood by others.
    "numeric", # connectedness - I feel connected to my community.
    "numeric", # baseline wellbeing - I have felt calm and at peace.
    "numeric", # baseline wellbeing - I have felt irritable or easily frustrated.
    "numeric", # baseline wellbeing - I have experienced feelings of loneliness.
    "numeric", # baseline wellbeing - I have felt down, depressed, or hopeless.
    "numeric", # baseline wellbeing - I have felt anxious or "on edge".
    "numeric", # baseline wellbeing - I have experienced physical pain or discomfort.
    "numeric", # baseline wellbeing - I have had difficulty concentrating or focusing.
    "numeric", # baseline wellbeing - I have had difficulty sleeping.
    "numeric", # baseline wellbeing - I have found it easy to make decisions.
    "numeric", # baseline wellbeing - I have felt motivated to accomplish daily tasks.
    "numeric", # baseline wellbeing - I have felt a sense of purpose or meaning in life.
    "numeric", # baseline wellbeing - I have experienced moments of awe or transcendence.
    "numeric", # baseline wellbeing - I have been overall satisfied with my life.
    "numeric", # baseline wellbeing - I have felt supported by my social network.
    "numeric", # baseline wellbeing - I have experienced conflict or tension in my relationships.
    "skip", # IMPORTANT: Before clicking Submit… Participant ID: [Field-…]
    "skip", # Score
    "text" # Participant ID
  ),
  skip = 1
) %>%
  mutate(
    timepoint = "1-Month"
  ) %>%
  rename(
    pid_auto = "Participant ID",
    pid_manual = "What is the Participant ID you were given at the end of your first questionnaire? - My Participant ID is... - Text",
    omo_response_id = `Response ID`,
    omo_recorded_date = `Recorded Date`,
    omo_greatest_emotion_txt = `greatest emotion`,
    omo_of_1_relaxed_and_at_peace = `overall feeling - Relaxed and at peace`,
    omo_of_2_connected_to_self = `overall feeling - Connected to myself`,
    omo_of_3_connected_to_others = `overall feeling - Connected to others`,
    omo_of_4_emotionally_balanced = `overall feeling - Emotionally balanced`,
    omo_of_5_insightful = `overall feeling - Insightful about my life or challenges`,
    omo_of_6_confused = `overall feeling - Confused and disoriented`,
    `omo_nsc_1 - I feel emotionally connected to the natural world.` =
      "connectedness - I feel emotionally connected to the natural world.",
    `omo_nsc_2 - I view humans as separate from nature.` =
      "connectedness - I view humans as separate from nature.",
    `omo_nsc_3 - I notice and appreciate the small details of nature.` =
      "connectedness - I notice and appreciate the small details of nature (e.g., plants, animals, weather).",
    `omo_nsc_4 - I feel a sense of awe and wonder when in nature.` =
      "connectedness - I feel a sense of awe and wonder when in nature.",
    `omo_nsc_5 - I feel motivated to make choices that are more environmentally friendly.` =
      "connectedness - I feel motivated to make choices that are more environmentally friendly (e.g., recycling, conserving energy).",
    `omo_nsc_6 - I feel a strong spiritual bond with the Earth and its living beings.` =
      "connectedness - I feel a strong spiritual bond with the Earth and its living beings.",
    `omo_nsc_7 - I feel connected to others.` =
      "connectedness - I feel connected to others.",
    `omo_nsc_8 - I feel disconnected from others.` =
      "connectedness - I feel disconnected from others.",
    `omo_nsc_9 - I feel like I belong.` =
      "connectedness - I feel like I belong.",
    `omo_nsc_10 - I feel understood by others.` =
      "connectedness - I feel understood by others.",
    `omo_nsc_11 - I feel connected to my community.` =
      "connectedness - I feel connected to my community.",
    `omo_wb_1 - I have felt calm and at peace.` =
      "baseline wellbeing - I have felt calm and at peace.",
    `omo_wb_2 - I have felt irritable or easily frustrated.` =
      "baseline wellbeing - I have felt irritable or easily frustrated.",
    `omo_wb_3 - I have experienced feelings of loneliness.` =
      "baseline wellbeing - I have experienced feelings of loneliness.",
    `omo_wb_4 - I have felt down, depressed, or hopeless.` =
      "baseline wellbeing - I have felt down, depressed, or hopeless.",
    `omo_wb_5 - I have felt anxious or \"on edge\".` =
      "baseline wellbeing - I have felt anxious or \"on edge\".",
    `omo_wb_6 - I have experienced physical pain or discomfort.` =
      "baseline wellbeing - I have experienced physical pain or discomfort.",
    `omo_wb_7 - I have had difficulty concentrating or focusing.` =
      "baseline wellbeing - I have had difficulty concentrating or focusing.",
    `omo_wb_8 - I have had difficulty sleeping.` =
      "baseline wellbeing - I have had difficulty sleeping.",
    `omo_wb_9 - I have found it easy to make decisions.` =
      "baseline wellbeing - I have found it easy to make decisions.",
    `omo_wb_10 - I have felt motivated to accomplish daily tasks.` =
      "baseline wellbeing - I have felt motivated to accomplish daily tasks.",
    `omo_wb_11 - I have felt a sense of purpose or meaning in life.` =
      "baseline wellbeing - I have felt a sense of purpose or meaning in life.",
    `omo_wb_12 - I have experienced moments of awe or transcendence.` =
      "baseline wellbeing - I have experienced moments of awe or transcendence.",
    `omo_wb_13 - I have been overall satisfied with my life.` =
      "baseline wellbeing - I have been overall satisfied with my life.",
    `omo_wb_14 - I have felt supported by my social network.` =
      "baseline wellbeing - I have felt supported by my social network.",
    `omo_wb_15 - I have experienced conflict or tension in my relationships.` =
      "baseline wellbeing - I have experienced conflict or tension in my relationships."
  ) %>%
  relocate(c(pid_auto, pid_manual, omo_response_id, omo_recorded_date, timepoint), .before = 1) %>%
  filter(Finished != 0) %>%
  select(-Finished)

# ---- pid coalescing before merge ---- ----

fu <- fu %>%
  mutate(
    pid_auto   = sub("\\.0$", "", pid_auto),
    pid_manual = sub("\\.0$", "", pid_manual),
    pid = coalesce(pid_manual, pid_auto)
  ) %>%
  relocate(c(pid_auto, pid_manual, pid, fu_response_id), .before = 1)

omo <- omo %>%
  mutate(
    pid_auto   = sub("\\.0$", "", pid_auto),
    pid_manual = sub("\\.0$", "", pid_manual),
    pid = coalesce(pid_manual, pid_auto)
  ) %>%
  relocate(c(pid_auto, pid_manual, pid, omo_response_id), .before = 1)

# ---- Manual matching ---- ----

# remove Michèle's test submission from baseline
basev <- basev %>%
  filter(bl_response_id != "R_6teomARAyFf1gge")

# fu, originally 46. manually matched one person, manually fixed another with a typo, one is dropped for not completing baseline.
fu <- fu %>%
  mutate( 
    pid = case_when(
      fu_response_id == "R_6e3J8ppiX1jtPs2" ~ "418328", # demographic match to baseline
      pid_manual == "566132" ~ "565132", # typo assumed to be same person because of retreat and only off by one number
      TRUE ~ pid
    )
  ) %>%
  select(-pid_auto, -pid_manual)

# omo, manually matched 3 ppl and will drop one for not completing baseline
omo <- omo %>%
  mutate(
    pid = case_when(
      omo_response_id == "R_3sTqbmgs0GDHhA0" ~ "391644", # demographic match to baseline
      omo_response_id == "R_407wj0qqBiCHFbI" ~ "774750", # demographic match to baseline
      omo_response_id == "R_8IKuKnBB3V9ZQm1" ~ "495946", # demographic match to baseline
      TRUE ~ pid
    )
  ) %>%
  select(-pid_auto, -pid_manual)
```

```{r audit for completion, eval=FALSE, include=FALSE}
# Who’s in Follow-up but not Baseline?
lost_fu <- fu %>% 
  anti_join(basev, by = "pid")

# Who’s in One-Month but not Baseline?
lost_omo <- omo %>%
  anti_join(basev, by = "pid")

# Combine
lost <- bind_rows(
  lost_fu  %>% mutate(Source = "Follow-up"),
  lost_omo %>% mutate(Source = "One-month")
)

# View the offending rows
lost %>% 
  select(pid, Source) %>% 
  distinct()
```

```{r baseline merge, include=FALSE}
# Join the baselines

baseline <- basev %>%
  left_join(
    basetxt,
    by     = c("bl_response_id" = "Response ID"),
    suffix = c("", "_txt")
  ) %>%
  relocate(
    
  # 1) IDs and timing
  pid,
  bl_response_id,
  bl_recorded_date,
  retreat,
  timepoint,
  
  # 2) Demographics
  age,
  gender,
  education,
  language,
  country_origin,
  country_residence,
  religion,
  
  
  # 4) Baseline experiences & intentions
  prev_ayahuasca,
  practices,
  intention,
  intent_mental,
  intent_physical,
  intent_other,
  bl_conn_def,
  
  .before = everything()
)
```

```{r demographics factoring, include=FALSE}

baseline <- baseline %>%
  mutate(
    # Standardize Country of Origin
    country_origin = case_when(
      country_origin %in% c("Brasil", "Brazil")                                ~ "Brazil",
      country_origin %in% c("United States", "United States of America", "USA")~ "United States",
      country_origin %in% c("Britain", "England")                              ~ "United Kingdom",
      country_origin == "México"                                               ~ "Mexico",
      country_origin == "France and Lebanon"                                   ~ "France",
      TRUE                                                                      ~ country_origin
    ),
    
    # Standardize Country of Residence
    country_residence = case_when(
      country_residence %in% c("Brasil", "Brazil")                             ~ "Brazil",
      country_residence %in% c("United States", "United States of America", "USA") ~ "United States",
      country_residence %in% c("Britain", "England", "Wales", "UK/India")      ~ "United Kingdom",
      TRUE                                                                      ~ country_residence
    ),

    # Retreats in chronological order
    retreat = factor(
      retreat,
      levels = c("Ninawa", "Colibri Festival", "Tumã", "Nixiwaka")
    ),
    
    gender = factor(gender),

    education = factor(
      education,
      levels = c(
        "Primary (Elementary)",
        "Upper Secondary (High)",
        "Vocational",
        "Bachelor's",
        "Master's",
        "Doctoral"
      ),
      ordered = TRUE
    ),

    # reorder & drop unused country levels
    country_origin    = fct_infreq(country_origin),
    country_residence = fct_infreq(country_residence),

    # reorder & drop unused intention levels
    intention = fct_infreq(intention),
    intention = fct_drop(intention),

    # Past ayahuasca experience (renamed prev_ayahuasca)
    prev_ayahuasca = factor(
      prev_ayahuasca,
      levels = c(
        "Never",
        "One time",
        "2–5 times",
        "6–10 times",
        "More than 10 times"
      ),
      ordered = TRUE
    ),
  
    # binary flag for spiritual intent
    spiritual_intent = if_else(intention == "Spiritual growth", 1L, 0L),

    # high vs low group on ayahuasca experience
    experience_group = fct_collapse(
      prev_ayahuasca,
      Low  = c("Never", "One time", "2–5 times"),
      High = c("6–10 times", "More than 10 times")
    )
  )
```

```{r all_wide join, include=FALSE}
# colnames(baseline)
# colnames(fu)
# colnames(omo)

all_wide <- baseline %>%
  left_join(fu, by = "pid", suffix = c("", ".fu")) %>%
  left_join(omo, by = "pid", suffix = c("", ".omo")) %>%
  select(-c(timepoint.omo, timepoint.fu))
```

```{=html}
<!-- # Abstract
[draft abstract]

# Introduction
[draft literature review]
-->
```

# Methods

This pilot study administered **three questionnaires**: *Baseline*, *Follow-up*, and *One-Month*, to participants at **four ceremonial ayahuasca retreats** held at **Aldeia Akasha** in Petrópolis, Rio de Janeiro, Brazil, between January and April 2025. Each retreat included **three ayahuasca ceremonies** conducted over the course of one week.

Participants completed the **Baseline questionnaire electronically** via Qualtrics at the start of the retreat, **prior to the first ceremony**. Informed consent was obtained through an affirmative response to the consent item included at the beginning of the questionnaire. Due to the small scale and programmatic nature of the study, it did not undergo formal IRB review.

The Baseline questionnaire collected:

-   **Demographic information** (e.g., age, gender, highest education level, country of origin and residence, primary language, religious affiliation, and prior spiritual/healing practices)

-   **Previous ayahuasca experience**

-   **Primary intention** for attending the retreat

To accommodate an **internationally diverse audience**, the education question included examples from various systems (e.g., *Ensino Fundamental*, *Lycée*, *Grundschule*), and questionnaires were offered in both **English and Portuguese**.

Participants also completed a section on **recent wellbeing over the past two weeks** (e.g., calmness, irritability, anxiety, depressive symptoms, and life satisfaction) in both the **Baseline** and **One-Month** questionnaires. A section measuring **nature and social connectedness** appeared in **all three** questionnaires.

All items used **5-point Likert scales** (e.g., from “Strongly disagree” to “Strongly agree” or “Not at all important” to “Extremely important”). Items were presented in randomized order within each section. Participants received an anonymous Participant ID for longitudinal tracking.

Immediately after the retreat, participants completed the **Follow-up questionnaire**, which repeated the **nature and social connectedness** measures and added items on:

-   **Emotional and cognitive effects** (e.g., peace, connection, emotional balance, and personal insights)

-   **Ceremonial elements** (e.g., presence of shamans, music, rituals, and setting)

-   **Primary post-retreat emotions** (open-ended responses)

The **One-Month questionnaire** reassessed **wellbeing**, **nature and social connectedness**, and **mental health status**, using the same items as earlier questionnaires for longitudinal comparison.

All analyses and visualizations were completed using RStudio Version 2025.05.0+496 from June to July 2025.

We gratefully acknowledge the generous support of the **Guardians and support staff of Aldeia Akasha** ([link](https://uniretreats.org/akasha-retreat-center/)), without whose openness, collaboration, and hospitality this pilot study would not have been possible.

# Results

## Participant Characteristics

The *Baseline* sample of 62 participants reflects a relatively mature, well-educated, and internationally diverse group. **Most were women (58.1%)** in their early forties on average, and over half held at least a bachelor's degree. Geographically, the largest groups came from **Brazil, the United Kingdom, and the United States**, consistent with Aldeia Akasha's location in Rio de Janeiro, Brazil, also popular with international audiences.

**Table 1. Baseline participant characteristics by retreat**

```{r table1, echo=FALSE}
## Table1 only from baseline retreat with matched participants

table1_selection <- baseline %>%
  select(
    retreat,
    gender,
    age,
    prev_ayahuasca,
    intention,
    education,
    country_origin,
    country_residence
  )

# Attach “label” attributes:
label(baseline$gender)            <- "Gender"
label(baseline$education)         <- "Highest education level"
label(baseline$country_residence) <- "Country of residence"
label(baseline$intention)         <- "Ceremony intention"
label(baseline$prev_ayahuasca)    <- "Times taken ayahuasca"
label(baseline$age)               <- "Age (years)"

table1(
  ~ gender
  + age
  + education
  + prev_ayahuasca
  + intention
  + country_residence  | retreat,
  data    = baseline,
  overall = "Total",
)

```

### Questionnaire Completion

**Only 39% of participants completed all three questionnaires**, underscoring the challenges of long-term follow-up. While 62 individuals enrolled at baseline, just 24 made it to the final one-month survey. Retention varied considerably across retreats: **Tumã's retreat had no participants complete the final questionnaire, while Ninawa's retreat retained over half of its *Baseline* participants (52%)** through to the end.

Retention at *Follow-up* was notably higher at **73% overall**. Nixiwaka's retreat stood out with the highest follow-up retention at 89%, likely attributable to this group's increased contact with the research team.

**Table 2: Questionnaire completion by Retreat**

```{r table2 completion, echo=FALSE}
# 1) Define the desired order
ordered_levels <- c("Ninawa", "Colibri Festival", "Tumã", "Nixiwaka", "Total")

# 2) Compute raw counts
table2 <- baseline %>%
  group_by(retreat) %>%
  summarize(
    N_baseline = n_distinct(pid),
    N_fu       = sum(pid %in% fu$pid),
    N_omo      = sum(pid %in% omo$pid),
    .groups    = "drop"
  )

# 3) Totals
total_baseline <- sum(table2$N_baseline)
total_fu       <- sum(table2$N_fu)
total_omo      <- sum(table2$N_omo)

# 4) Build display rows for non-Total
display <- table2 %>%
  filter(retreat != "Total") %>%
  mutate(
    Baseline  = as.character(N_baseline),
    `Follow-up` = sprintf(
      "%d (%d%%)", N_fu,
      round(100 * N_fu / N_baseline)
    ),
    `1-Mo`    = sprintf(
      "%d (%d%%)", N_omo,
      round(100 * N_omo / N_baseline)
    )
  ) %>%
  select(retreat, Baseline, `Follow-up`, `1-Mo`)

# 5) Append Total row with counts only
display <- bind_rows(
  display,
  tibble(
    retreat      = "Total",
    Baseline     = as.character(total_baseline),
    `Follow-up`  = sprintf(
      "%d (%d%%)",
      total_fu,
      round(100 * total_fu / total_baseline)
    ),
    `1-Mo`       = sprintf(
      "%d (%d%%)",
      total_omo,
      round(100 * total_omo / total_baseline)
    )
  )
)

# 6) Order and render
display %>%
  mutate(retreat = factor(retreat, levels = ordered_levels)) %>%
  arrange(retreat) %>%
  kable(
    format    = "html",
    align     = c("l", "r", "r", "r"),
    escape    = FALSE,
    col.names = c(
      "Retreat",
      "Baseline (<i>n</i>)",
      "Follow-up (<i>n</i>, % retention)",
      "1-Mo (<i>n</i>, % retention)"
    )
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width        = TRUE,
    position          = "left"
  ) %>%
  row_spec(nrow(display), bold = TRUE)

```

```{r prep timepoint comparison, echo=FALSE}
# 1) Bind all three waves with new prefixes (nsc_ and wb_)
all_raw <- bind_rows(
  baseline %>%
    mutate(timepoint = "Baseline") %>%
    select(
      pid, retreat, timepoint, prev_ayahuasca, intention,
      starts_with("bl_nsc_"), starts_with("bl_wb_")
    ),
  fu %>%
    mutate(timepoint = "Follow-up") %>%
    select(
      pid, timepoint,
      starts_with("fu_nsc_"), starts_with("fu_wb_")
    ),
  omo %>%
    mutate(timepoint = "One-month") %>%
    select(
      pid, timepoint,
      starts_with("omo_nsc_"), starts_with("omo_wb_")
    )
)

# 2) Pivot long
all_long <- all_raw %>%
  pivot_longer(
    cols      = -c(pid, timepoint, retreat, prev_ayahuasca, intention),
    names_to  = "Question_raw",
    values_to = "response"
  )

# 3) Split NATSC vs wellbeing by prefix
all_nat        <- all_long %>% filter(str_detect(Question_raw, "^.*_nsc_"))
all_wellbeing  <- all_long %>% filter(str_detect(Question_raw, "^.*_wb_"))


# define which questions to invert
inv_nat <- c(
  "I view humans as separate from nature.",
  "I feel disconnected from others."
)
inv_wb <- c(
  "I have experienced feelings of loneliness.",
  "I have felt anxious or \"on edge\".",
  "I have had difficulty concentrating or focusing.",
  "I have experienced conflict or tension in my relationships.",
  "I have felt irritable or easily frustrated.",
  "I have felt down, depressed, or hopeless.",
  "I have experienced physical pain or discomfort.",
  "I have had difficulty sleeping."
)

# 5) Clean NATSC
cleaned_nat <- all_nat %>%
  mutate(
    Question = str_remove(Question_raw, "^.*_nsc_\\d+\\s*-\\s*"),
    # invert where needed
    response = if_else(Question %in% inv_nat, 6L - response, response),
    Question = if_else(
      Question %in% inv_nat,
      str_c(Question, " (inv.)"),
      Question
    )
  )

# 6) Clean wellbeing
cleaned_wb <- all_wellbeing %>%
  mutate(
    Question = Question_raw %>%
      str_remove("^.*_wb_") %>%            # drop everything through "_wb_"
      str_remove("^\\d+\\s*-\\s*"),        # drop leading "1 - " (or "2 - ", etc)
    response = if_else(Question %in% inv_wb, 6L - response, response),
    Question = if_else(Question %in% inv_wb,
      str_c(Question, " (inv.)"),
      Question
    )
  )

# Commented this out because it isn't used anywhere
# # 7) Full cleaned for combined use
# cleaned_all <- bind_rows(
#   cleaned_nat   %>% select(pid, retreat, timepoint, Question, response),
#   cleaned_wb    %>% select(pid, retreat, timepoint, Question, response)
# )

# 8) Summaries
summary_nat <- cleaned_nat %>%
  group_by(Question, timepoint) %>%
  summarize(mean_response = mean(response, na.rm = TRUE), .groups = "drop")

summary_wb <- cleaned_wb %>%
  group_by(Question, timepoint) %>%
  summarize(mean_response = mean(response, na.rm = TRUE), .groups = "drop") %>%
  filter(timepoint != "Follow-up")
```

### Who's Taken Ayahuasca Before?

Chart 1 shows that prior ayahuasca experience among participants was polarized at two extremes. Over a third of participants had taken ayahuasca **more than 10 times (35%)**, while the next largest group **(21%) had never taken it at all**. This bimodal distribution suggests the retreat appealed to both committed users seeking deeper work and newcomers drawn to its promise. The intermediate groups were slightly smaller (11–18%), hinting at a possible threshold beyond which individuals either commit more fully or opt out. The sample thus reflects a dynamic mix of ayahuasca-naive and highly experienced participants, with smaller proportions in between.

**Chart 1: How Many Times Have Participants Taken Ayahuasca?**

```{r ayahuasca frequency, echo=FALSE, fig.align='left', fig.width=9}

freq <- baseline %>%
  count(prev_ayahuasca) %>%
  arrange(match(prev_ayahuasca,
    c("Never","One time","2–5 times","6–10 times","More than 10 times")
  )) %>%
  mutate(
    pct = n / sum(n),
    label = paste0(n, " (", scales::percent(pct, accuracy = 1), ")"),
    prev_ayahuasca = factor(
      prev_ayahuasca,
      levels = c("Never","One time","2–5 times","6–10 times","More than 10 times")
    )
  )

ggplot(freq, aes(x = prev_ayahuasca, y = n)) +
  geom_col(fill = "#95d5b2", width = 0.7) +
  geom_text(
    aes(label = label),
    vjust = -0.5,
    color = "black",
    size = 5
  ) +
  scale_y_continuous(expand = expansion(c(0, .1))) +
  labs(
    x = "Times Taken Ayahuasca",
    y = "# Participants"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title.x = element_text(margin = margin(t = 12)),
    axis.title.y = element_text(margin = margin(r = 12)),
    axis.text          = element_text(size = 13)
  )

```

### Why Are Participants Taking Ayahuasca?

Most participants reported attending for spiritual growth (50%), followed by mental health and career motivations. However, when broken down by past ayahuasa experience, a more nuanced pattern emerged.

**Chart 2A: What are Participants' Intentions?**

```{r 2A prev overall, echo=FALSE, fig.align='left'}

# ---- Define consistent levels and colors ----

# Define consistent levels for intentions
intent_levels <- c(
  "Spiritual growth", 
  "Mental health", 
  "Career", 
  "Other", 
  "Physical health", 
  "Curiosity"
)

# Assign consistent colors manually
intent_colors <- c(
  "Spiritual growth" = "#66c2a5",  # green
  "Mental health"    = "#fc8d62",  # orange
  "Career"           = "#8da0cb",  # blue
  "Other"            = "#e78ac3",  # pink
  "Physical health"  = "#ffd92f",  # yellow
  "Curiosity"        = "#a6d854"   # lime green
)

# ---- chart ----

overall_intent <- baseline %>%
  count(intention) %>%
  mutate(
    prop = n / sum(n),
    label = paste0(n, " (", scales::percent(prop, accuracy = 1), ")"),
    intention = fct_reorder(intention, prop)
  )

ggplot(overall_intent, aes(x = intention, y = prop, fill = intention)) +
  geom_col(width = 0.7, color = "white") +
  geom_text(aes(label = label), hjust = -0.1, size = 4) +
  scale_y_continuous(labels = scales::percent_format(), expand = expansion(mult = c(0, 0.2))) +
  scale_fill_manual(values = intent_colors, guide = "none") +
  coord_flip() +
  labs(
    x = "Intention",
    y = "Proportion of Participants",
    caption = expression(italic(n) == 62)
  ) +
  theme_light(base_size = 13) +
  theme(panel.grid.minor = element_blank())

```

Participants' intentions reveal a shift in orientation with increasing ayahuasca experience. **Spiritual growth** dominated among those with 6 or more previous sessions, and especially in the most experienced group, among whom 68% reported it as their main reason for attending. In contrast, participants with moderate experience (2–5 sessions) were more likely to cite **mental health** as their primary intention (57%), suggesting that early encounters with ayahuasca may be motivated by psychological growth, with spiritual growth emerging as an intention as experience deepens. Less frequent intentions such as curiosity or physical health appeared sporadically across groups, while career-related goals remained steady, possibly reflecting a distinct motivational category. Together, these patterns imply an evolving relationship to ayahuasca shaped by accumulated exposure.

**Chart 2B: Did Participants' Intentions Change Based on Their Previous Experience?**

```{r 2B prev by intention, echo=FALSE, fig.align='left'}
# Prepare counts + proportions
intention_freq <- baseline %>%
  count(prev_ayahuasca, intention) %>%
  group_by(prev_ayahuasca) %>%
  mutate(
    prop = n / sum(n),
    text = paste0(
      "Experience: ", prev_ayahuasca, "<br>",
      "Intention:  ", intention,      "<br>",
      "Count:      ", n,              "<br>",
      "Percent:    ", scales::percent(prop, 1)
    )
  ) %>%
  ungroup() %>%
  mutate(
    prev_ayahuasca = factor(
      prev_ayahuasca,
      levels = c("Never","One time","2–5 times","6–10 times","More than 10 times")
    )
  )

ggplot(intention_freq, aes(x = prev_ayahuasca, y = prop, fill = intention)) +
  geom_col(color = "white", position = "fill", width = 0.7) +
  geom_text(
    aes(label = ifelse(prop >= 0.08, scales::percent(prop, 1), "")),
    position = position_fill(vjust = 0.5),
    color = "black",
    size = 3.5
  ) +
  scale_y_continuous(labels = scales::percent_format(), expand = expansion(0)) +
  scale_fill_manual(values = intent_colors) +
  coord_flip() +
  labs(
    x = "Times Taken Ayahuasca",
    y = "Proportion of Intentions",
    fill = "Ceremony Intention",
    caption = expression(italic(n) == 62)
  ) +
  theme_light(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```

Indeed, participants who had attended 6 or more previous ceremonies (*n* = 31) were significantly more likely to report spiritual growth as their primary intention compared to those with less experience. In a logistic regression adjusting for retreat, participants in the high-experience group had more than twice the odds of reporting a spiritual intention (OR = 2.53, 95% CI: 1.16 to 5.82, *p* = 0.024\*). This statistically significant association suggests that spiritual intentions may become more prominent with continued engagement in ayahuasca ceremonies, and could merit further exploration in future studies.

```{r logistic experience x intention, include=FALSE, echo=FALSE}
# tbl <- table(baseline$experience_group,
#              baseline$spiritual_intent)
# 
# fisher.test(tbl, alternative = "two.sided")

fisherglm = glm(spiritual_intent ~ experience_group + retreat, data = baseline, family = binomial)
summary(fisherglm)

tidy(fisherglm,
     conf.int     = TRUE,     # ask for confidence intervals
     exponentiate = TRUE      # return exp(estimate) instead of the log‐odds
)
```

## Ceremonial Setting

In the *Follow-up* questionnaire, participants were asked to reflect on the ceremonial contexts of their ayahuasca ceremonies. This section aimed to assess the perceived importance of various environmental and social elements within the ceremony. Participants rated eight aspects, including the presence of the shaman or *pajé* (e.g., Nixiwaka, Ninawa, Putanny), ritual elements (e.g., music, prayers), and contextual factors (e.g., group setting, physical space, sharing circles, and seat partners), on a 5-point Likert scale ranging from "Not at all important" to "Extremely important." These items were designed to capture how components of the setting may have contributed to participants’ subjective experience.

Participants overwhelmingly agreed that **music** was the single most important part of the ceremony, assigning it an average importance of 4.91 out of 5 and with 97.8% of respondents rating it "very" or "extremely" important. The physical environment came next (M = 4.65; 95.7% top-2-box), closely followed by the presence of the shaman, the rituals and prayers, and the group setting itself. In contrast, the sharing circle was seen as somewhat less essential (M = 4.04; 76.1%), and one's immediate neighbors was least critical of all (M = 3.63; 52.5%). These results suggest that while most aspects of the ceremonial context were viewed as highly important, **participants place particularly high value on music and relatively less on one's immediate neighbors**.

**Chart 4: Which Aspects of the Ceremony Were Most Important to Participants?**

```{r ceremonial setting clean, echo=FALSE, fig.align='left', fig.width=10}
# 1. Pivot all your ceremony aspects (including both shaman vars) long
ceremonies_long <- fu %>%
  select(
    pid,
    fu_setting_Presence_Shaman,
    fu_setting_Presence_Putany,
    fu_setting_Music,
    fu_setting_Rituals_Prayers,
    fu_setting_Group_Setting,
    fu_setting_Physical_Environment,
    fu_setting_Sharing_Circle,
    fu_setting_Immediate_Neighbors
  ) %>%
  pivot_longer(
    cols      = -pid,
    names_to  = "Raw_Aspect",
    values_to = "Rating"
  ) %>%
  filter(!is.na(Rating)) %>%
  mutate(
    Aspect = case_when(
      Raw_Aspect %in% c("fu_setting_Presence_Shaman", "fu_setting_Presence_Putany") ~ "Shaman presence",
      Raw_Aspect == "fu_setting_Music"            ~ "Music",
      Raw_Aspect == "fu_setting_Rituals_Prayers"  ~ "Rituals & prayers",
      Raw_Aspect == "fu_setting_Group_Setting"    ~ "Group setting",
      Raw_Aspect == "fu_setting_Physical_Environment" ~ "Physical environment",
      Raw_Aspect == "fu_setting_Sharing_Circle"   ~ "Sharing circle",
      Raw_Aspect == "fu_setting_Immediate_Neighbors" ~ "Immediate neighbors",
      TRUE                                         ~ Raw_Aspect
    ),
    Rating = factor(
      Rating,
      levels = 1:5,
      labels = c(
        "Not at all important",
        "Slightly important",
        "Moderately important",
        "Very important",
        "Extremely important"
      ),
      ordered = TRUE
    )
  )

# calculate
mean_ratings <- ceremonies_long %>% 
  mutate(r = as.numeric(Rating)) %>% 
  group_by(Aspect) %>% 
  summarize(mean_r = mean(r, na.rm=TRUE)) %>% 
  arrange(mean_r) %>% 
  mutate(Aspect = fct_inorder(Aspect))

summary_tbl <- ceremonies_long %>%
  mutate(r = as.numeric(Rating)) %>%
  group_by(Aspect) %>%
  summarize(
    Mean       = round(mean(r),2),
    Top2BoxPct = percent(mean(Rating %in% c("Very important","Extremely important")), 0.1)
  ) %>%
  arrange(desc(Mean)) 

knitr::kable(summary_tbl, format="html", align="lrr", escape = FALSE, col.names = c("Ceremonial Aspect", "Mean (1 – 5)", "% Very/Extremely<br>Important")) %>%
  kableExtra::kable_styling(full_width=FALSE) %>%
  kableExtra::row_spec(0, bold=TRUE) %>%
  #kableExtra::column_spec(2, background = "#EFF3FF") %>%  # shade the Mean column
  kableExtra::footnote(
    general = "<i>n</i> = 46, including one participant who did not finish <i>Baseline</i> but was included here for descriptive purposes",
    escape = FALSE)


```

Participants were also asked to describe the ceremonial setting in one word. Common themes included feelings of *safety*, *connection*, *being held*, and a sense of *belonging*, with repeated mentions of terms like *safe*, *held*, *oneness*/*communal*. Many described the setting as *sacred*, *powerful*, and even *cosmic*, using terms such as magical, ecstasy, oneness, and celebração.

This description suggests that participants experienced the ceremonial setting as not just a physical or logistical arrangement, but also **emotionally and spiritually charged**. Participants' language reflect that the setting fostered a profound sense of emotional security and relational trust. Simultaneously, words like *sacred*, *cosmic*, and *ecstasy* point to an **expanded or transcendent experience**, where the environment supported states of awe, unity, and transformation. Overall, participants appear to have experienced the setting not only as structured and supportive, but as a container for connection and meaning.

## Nature & Social Connectedness

One of our intentions in running this pilot study was to determine to what extent ayahuasca affected people's sense of connection to nature, and connection to one another. \[write about the scale\]

Across the three questionnaires, participants reported consistently high levels of connectedness to nature. Responses to several questions showed steady increases over time, **particularly those related to social connection**. For example, average ratings for “I feel connected to others” and “I feel understood by others” rose significantly from *Baseline* to *Follow-up,* and continued increasing by *One-month*. In contrast, items like “I feel a strong spiritual bond with the Earth and its living beings” or “I feel emotionally connected to the natural world” remained high, but unchanged after the ceremonies.

Notably, the item “I feel disconnected from others” (inverted) showed a clear upward trend, indicating **reduced feelings of social disconnection following the ceremony**.

Chart 3 shows mean responses to each **Nature and Social Connectedness** item at three timepoints, where 5 = Strongly Agree, 4 = Somewhat Agree, 3 = Neither agree nor disagree, 2 = Somewhat Disagree, and 1 = Strongly Disagree.

**Chart 3: How Participants' Nature and Social Connectedness Evolved After Retreat**

```{r chart: naturesocial across timepoints, echo=FALSE, fig.width=16, fig.height=20}
labeltextsize <- 7

### nat
ggplot(summary_nat, aes(x = timepoint, y = mean_response, group = 1)) +
  geom_line(color = "gray40") +
  geom_point(aes(color = timepoint), size = 3) +
  geom_text(
    aes(label = sprintf("%.2f", mean_response)),
    color = "black",
    vjust = 2.0,
    size = labeltextsize
  ) +
  geom_text(
  aes(label = timepoint),
  color = "black",
  vjust = 4.5,
  size = labeltextsize - 2
  ) +
  facet_wrap(
    ~ Question,
    nrow   = 5,
    scales = "fixed",
    labeller = label_wrap_gen(width = 30)
  ) +
  scale_y_continuous(limits = c(1, 5), breaks = 1:5) +
  scale_color_manual(
    values = c(
      "Baseline"  = "#ace3a6",
      "Follow-up" = "#2eb51f",
      "One-month" = "#116b07"
    )
  ) +
  theme_minimal(base_size = 20) +
  theme(
    panel.spacing    = unit(0.5, "lines"),
    panel.border     = element_rect(color = "black", linewidth = 0.6, fill = NA),
    strip.text       = element_text(face = "bold"),
    axis.text.x      = element_blank(),
    panel.grid.major = element_line(color = "gray90"),
    legend.position  = "none"
  ) +
  labs(
    x = NULL,
    y = "Mean response (1–5)"
  )

```

These patterns were further supported by linear mixed model results that account for longtitudinal measures, and are adjusted for participants who did not complete all three timepoints. Statistically significant improvements were found between *Baseline* and both *Follow-up* and *One-month* questionnaires for several items, especially in the **social connectedness** domain. The strongest effects were seen in **reduced social disconnection** (Estimate = –1.04, *p* \< .001\*\*\*) and increases in feeling understood (Estimate = –1.01, *p* \< .001\*\*\*). These are highly statistically significant findings, in addition to being large effect sizes (1.0, in this case, refers to a one-point improvement between, e.g., Somewhat Agree and Strongly Agree).

In contrast, nature-related items, while high across the board, did not show statistically significant changes over time. These findings suggest that **participants already felt strongly connected to nature**—and while ayahuasca did not meaningfully affect this, the ceremony may have had a greater impact on **improving social and emotional connection with others**.

------------------------------------------------------------------------

**Table 4: Mixed Models on Nature and Social Connectedness**

Table 4 summarizes paired comparisons between *Baseline* and both subsequent questionnaires (*Follow-up* and *One-month*) across 11 questions on Nature and Social connectedness. Linear mixed models with random intercepts were used to estimate mean differences while accounting for repeated measures and including participants who did not complete all three surveys.

Statistically significant results are color-coded: \*\*\* p \< .001, (teal) \*\* p \< .01, (green) \* p \< .05.

```{r table: nature-social lmm, echo=FALSE}
results_nat_complete <- cleaned_nat %>%
  # only keep pid/timepoint/Question/response
  select(pid, timepoint, Question, response) %>%
  # for reproducibility, ensure timepoint is a factor in the right order
  mutate(timepoint = factor(timepoint, levels = c("Baseline", "Follow-up", "One-month"))) %>%
  # nest by Question
  nest_by(Question) %>%
  mutate(
    # fit LMM
    model = list(lmer(response ~ timepoint + (1 | pid), data = data, REML = FALSE)),
    # emmeans + pairwise contrasts
    em = list(emmeans(model, ~timepoint)),
    contrasts = list(as.data.frame(contrast(em, method = "pairwise", adjust = "none")))
  ) %>%
  # unnest the contrasts
  select(Question, contrasts) %>%
  ungroup() %>%
  unnest(contrasts)

slim_results_nat <- results_nat_complete %>%
  mutate(
    contrast = str_replace(contrast, " - ", " to "),
    contrast = str_remove_all(contrast, "\\(|\\)"),
    Estimate = round(estimate, 2),
    SE       = round(SE, 2),
    DF       = round(df, 1),
    t_value  = round(t.ratio, 2),
    p_fmt    = case_when(
      p.value < 0.001 ~ "<.001",
      p.value < 0.01  ~ "<.01",
      TRUE            ~ sprintf("%.3f", p.value)
    ),
    stars = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    p_colored = case_when(
      p.value < 0.001 ~ cell_spec(paste0(p_fmt, stars), color = "white", background = "#5E4FA2", bold = TRUE),
      p.value < 0.01  ~ cell_spec(paste0(p_fmt, stars), color = "white", background = "#3288BD", bold = TRUE),
      p.value < 0.05  ~ cell_spec(paste0(p_fmt, stars), color = "black", background = "#ABDDA4"),
      TRUE            ~ paste0(p_fmt, stars)
    ),
    Estimate_colored = case_when(
      p.value < 0.001 ~ cell_spec(sprintf("%.2f", Estimate), color = "white", background = "#5E4FA2", bold = TRUE),
      p.value < 0.01  ~ cell_spec(sprintf("%.2f", Estimate), color = "white", background = "#3288BD", bold = TRUE),
      p.value < 0.05  ~ cell_spec(sprintf("%.2f", Estimate), color = "black", background = "#ABDDA4"),
      TRUE            ~ sprintf("%.2f", Estimate))
  ) %>%
  filter(contrast != "Follow-up to One-month") %>%
  select(
    Question,
    Contrast = contrast,
    Estimate = Estimate_colored,
    SE,
    DF,
    `p value` = p_colored
  )

k <- kable(
  slim_results_nat,
  format = "html",
  escape = FALSE,
  align = c("l", "l", "r", "r", "r", "r"),
  col.names = c("Question", "Timeline Pair", "Estimate", "SE", "DF", "p-value")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    position = "left"
  ) %>%
  footnote(
    general = "Significance codes: * p < .05; ** p < .01; *** p < .001"
  )

# font-size rearrangement loop
for (i in seq_len(ncol(slim_results_nat))) {
  k <- k %>% column_spec(i, extra_css = "font-size: 15px;")
}

k
```

## Recent Wellbeing

\[copy about this section\]

### Table 5: How Participants' Recent Wellbeing Evolved After Retreat

Mean response on each **Recent Wellbeing** section at two timepoints, where 5 = Strongly Agree and 1 = Strongly Disagree. Note that recent wellbeing Questions were not asked at *Follow-up*, only at *Baseline* and *One-month*.

```{r chart: wellbeing across timepoints, echo=FALSE, fig.width=16, fig.height=16}
## wellbeing

ggplot(summary_wb, aes(timepoint, mean_response, group = 1)) +
  geom_line(color = "gray40") +
  geom_point(aes(color = timepoint), size = 5) +
  geom_text(
    aes(label = sprintf("%.2f", mean_response)),
    color = "black",
    vjust = 2,
    size = labeltextsize
  ) +
  geom_text(
  aes(label = timepoint),
  color = "black",
  vjust = 4.5,
  size = labeltextsize - 2
  ) +
  facet_wrap(
    ~ Question,
    nrow   = 5,
    scales = "fixed",
    labeller = label_wrap_gen(width = 30)
  ) +
  scale_y_continuous(limits = c(1, 5), breaks = 1:5) +
  scale_color_manual(
    values = c(
      "Baseline"   = "#A6CEE3",
      "One-month"  = "#08306B"
    )
  ) +
  theme_minimal(base_size = 20) +
  theme(
    panel.spacing    = unit(0.5, "lines"),
    panel.border     = element_rect(color = "black", linewidth = 0.6, fill = NA),
    strip.text       = element_text(face = "bold"),
    axis.text.x      = element_blank(),
    panel.grid.major = element_line(color = "gray90"),
    legend.position  = "none"
  ) +
  labs(
    x = NULL,
    y = "Mean response (1–5)"
  )
```

### Table 6: Mixed Models on Recent Wellbeing

Participants showed **broad and statistically significant** improvements across emotional, existential, and cognitive domains from baseline to one month following the retreat. As shown in *Table 6*, participants reported **reduced symptoms of psychological distress**, including **loneliness** (β = -0.59, p \< .001\*\*\*) and anxiety (β = -0.49, p \< .01\*\*), depression (β = -0.44, p \< .01\*\*), difficulty concentrating (β = -0.41, p \< .01\*\*), and irritability (β = -0.37, p = .032\*). These estimates (β) represent decreases of roughly one-third to over half a scale point on the 5-point Likert scale used, corresponding to **moderate to large effect sizes.**

Positive shifts were also observed in life satisfaction (β = -0.49, p = .013\*), calmness and peace (β = -0.61, p \< .01\*\*), and ease making decisions (β = -0.50, p = .023\*). Participants also reported **more frequent experiences of awe or transcendence** (β = -0.80, p \< .001\*\*\*), a change close to a full scale point, suggesting a particularly strong impact in this domain.

Other outcomes including sense of purpose, physical discomfort, sleep quality, and perceived social support trended in a positive direction, but did not reach statistical significance. No change was observed in relationship conflict. Taken together, these results suggest a broad and meaningful **improvement in psychological wellbeing, emotional regulation, and existential engagement** one month after the retreat.

```{r table: wellness lmm, echo=FALSE}
results_wb_complete <- cleaned_wb %>%
  # only keep pid/timepoint/Question/response
  select(pid, timepoint, Question, response) %>%
  # for reproducibility, ensure timepoint is a factor in the right order
  mutate(timepoint = factor(timepoint, levels = c("Baseline", "One-month"))) %>%
  # nest by Question
  nest_by(Question) %>%
  mutate(
    # fit LMM
    model = list(lmer(response ~ timepoint + (1 | pid), data = data, REML = FALSE)),
    # emmeans + pairwise contrasts
    em = list(emmeans(model, ~timepoint)),
    contrasts = list(as.data.frame(contrast(em, method = "pairwise", adjust = "none")))
  ) %>%
  # unnest the contrasts
  select(Question, contrasts) %>%
  ungroup() %>%
  unnest(contrasts)

slim_results_wb <- results_wb_complete %>%
  mutate(
    contrast = str_replace(contrast, " - ", " to "),
    contrast = str_remove_all(contrast, "\\(|\\)"),
    Estimate = round(estimate, 2),
    SE       = round(SE, 2),
    DF       = round(df, 1),
    t_value  = round(t.ratio, 2),
    p_fmt    = case_when(
      p.value < 0.001 ~ "<.001",
      p.value < 0.01  ~ "<.01",
      TRUE            ~ sprintf("%.3f", p.value)
    ),
    stars = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    p_colored = case_when(
      p.value < 0.001 ~ cell_spec(paste0(p_fmt, stars), color = "white", background = "#5E4FA2", bold = TRUE),
      p.value < 0.01  ~ cell_spec(paste0(p_fmt, stars), color = "white", background = "#3288BD", bold = TRUE),
      p.value < 0.05  ~ cell_spec(paste0(p_fmt, stars), color = "black", background = "#ABDDA4"),
      TRUE            ~ paste0(p_fmt, stars)
    ),
    Estimate_colored = case_when(
      p.value < 0.001 ~ cell_spec(sprintf("%.2f", Estimate), color = "white", background = "#5E4FA2", bold = TRUE),
      p.value < 0.01  ~ cell_spec(sprintf("%.2f", Estimate), color = "white", background = "#3288BD", bold = TRUE),
      p.value < 0.05  ~ cell_spec(sprintf("%.2f", Estimate), color = "black", background = "#ABDDA4"),
      TRUE            ~ sprintf("%.2f", Estimate)
    )
  ) %>%
  select(
    Question,
    Contrast = contrast,
    Estimate = Estimate_colored,
    SE,
    DF,
    `p value` = p_colored
  )

kable(
  slim_results_wb,
  format = "html",
  escape = FALSE, # <-- important to render HTML from cell_spec
  align = c("l", "l", "r", "r", "r", "r"),
  col.names = c("Question", "Timeline Pair", "Estimate", "SE", "DF", "p-value")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    position = "left"
  ) %>%
  footnote(
    general = "Significance codes: * p < .05; ** p < .01; *** p < .001"
  )

```

```{r qual exploration}
nrc <- get_sentiments("nrc")

# Combine definitions
conn_defs <- baseline %>%
  select(pid, bl_conn_def) %>%
  rename(pre = bl_conn_def) %>%
  left_join(select(fu, pid, fu_conn_def_txt), by = "pid") %>%
  rename(post = fu_conn_def_txt)

# Clean and tokenize
conn_defs_tokens <- conn_defs %>%
  pivot_longer(cols = c(pre, post), names_to = "timepoint", values_to = "text") %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word")

conn_defs_tokens %>%
  count(timepoint, word, sort = TRUE) %>%
  group_by(timepoint) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(word, n), y = n, fill = timepoint)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~timepoint, scales = "free") +
  coord_flip() +
  labs(title = "Most Common Words in Connection Definitions", x = "", y = "Count")

conn_defs_tokens %>%
  inner_join(get_sentiments("nrc")) %>%
  count(timepoint, sentiment, sort = TRUE)

sentiment_by_pid <- conn_defs_tokens %>%
  inner_join(nrc, by = "word") %>%
  count(pid, timepoint, sentiment) %>%
  pivot_wider(names_from = timepoint, values_from = n, values_fill = 0) %>%
  mutate(diff = post - pre)

conn_defs_tokens %>%
  inner_join(nrc, by = "word") %>%
  count(timepoint, sentiment, sort = TRUE) %>%
  group_by(timepoint) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

conn_defs_tokens %>%
  inner_join(nrc, by = "word") %>%
  count(timepoint, sentiment) %>%
  ggplot(aes(x = sentiment, y = n, fill = timepoint)) +
  geom_col(position = "dodge") +
  labs(
    title = "Sentiment Composition of Connection Definitions",
    x = "Sentiment", y = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r beep, message=FALSE, warning=FALSE, include=FALSE}
library(beepr)
beep()
```
